OptiFit: an improved method for fitting amplicon sequences to existing OTUs

### 2021-11-09 paper/paper.Rmd
Cherry-pick: Reduce abstract to 250 words

---
title: "OptiFit: an improved method for fitting amplicon sequences to existing OTUs"
date: '`r Sys.Date()`'
bibliography: references.bib
link-citations: true
output:
  pdf_document:
    includes:
      in_header: preamble.tex
      before_body: head.tex
    fig_caption: yes
  github_document:
    html_preview: false
csl: msphere.csl
fontsize: 12pt
geometry: margin=1.0in
params:
    snakemake: "none"
---

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
schtools::set_knitr_opts()
knitr::opts_knit$set(base.dir = here::here(), 
                     base.url = "../")
knitr::opts_chunk$set(fig.path = 'figures/',
                      dpi = 300)
```

```{r deps, include=FALSE, message = FALSE}
library(assertthat)
library(cowplot)
library(glue)
library(here)
library(knitr)
library(schtools)
library(tidyverse)
set.seed(20210624)
snakemake <- params$snakemake
if (is_character(snakemake)) {
  stop("You need to knit this R Markdown document from the Snakemake workflow")
}

@@

figures_meta <- yaml::read_yaml(here('paper', 'figures.yaml'))
figs <- function(fig_name, attribute) {
  value <- figures_meta[[fig_name]][[attribute]]
  if (attribute == 'cap') {
    value <- glue("\\label{fig:[fig_name]}[value]", .open='[', .close = ']')
  } else {
    value <- eval(parse(text = value))
  }
  return(value)
}

@@

Assigning amplicon sequences to operational taxonomic units (OTUs) is often an
important step in characterizing the composition of microbial communities across
large datasets.
OptiClust, a _de novo_ OTU clustering method, has been
shown to produce higher quality OTU assignments than other methods and at
comparable or faster speeds.
A notable difference between _de novo_ clustering and database-dependent
reference clustering methods is that OTU assignments from _de novo_ methods may
change when new sequences are added to a dataset.
However, in some cases one may wish to incorporate new samples into a previously
clustered dataset without performing clustering again on all sequences, such as
when comparing across datasets or deploying machine learning models where OTUs
are features.
Existing reference-based clustering methods produce consistent OTUs, but they
only consider the similarity of each query sequence to a single reference
sequence in an OTU, thus resulting in OTU assignments that are significantly
worse than those generated by _de novo_ methods.
To provide an efficient and robust method to fit amplicon sequence data to
existing OTUs, we developed the OptiFit algorithm.
Inspired by OptiClust, OptiFit considers the similarity of all pairs of reference and
query sequences in an OTU to produce OTUs of the best possible quality.
We tested OptiFit using four microbiome datasets with two different strategies:
by clustering to an external reference database or by splitting the dataset into a
reference and query set and clustering the query sequences to the reference set
after clustering it using OptiClust.
The result is an improved implementation of closed and open-reference clustering.
OptiFit produces OTUs of similar quality as OptiClust and at faster speeds when
using the split dataset strategy, although the OTU quality and processing speed
depends on the database chosen when using the external database strategy.
OptiFit provides a suitable option for users who require consistent OTU
assignments at the same quality afforded by _de novo_ clustering methods.

@@

```{r algorithm, fig.dim=figs('algorithm','dim'), fig.cap=figs('algorithm','cap')}
figs('algorithm', 'path') %>%
  tiff::readTIFF() %>%
  grid::grid.raster()
```

@@

```{r workflow, fig.dim=figs('workflow', 'dim'), fig.cap=figs('workflow','cap')}
figs('workflow', 'path') %>%
  tiff::readTIFF() %>%
  grid::grid.raster()
```

@@

```{r results_sum, fig.dim=figs('results_sum', 'dim'), fig.cap=figs('results_sum', 'cap')}
figs('results_sum', 'path') %>%
  tiff::readTIFF() %>%
  grid::grid.raster()
```

@@

```{r results_split, fig.dim=figs('results_split', 'dim'), fig.cap=figs('results_split', 'cap')}
figs('results_split', 'path') %>%
  tiff::readTIFF() %>%
  grid::grid.raster()
```
```{r, include = FALSE}
close_enough(cv_fit_split_mcc_human_simple, 0, tol = 10^-3)
assert_that(sec_fit_split_human_simple_1 > sec_fit_split_human_simple_9)
assert_that(frac_fit_split_human_simple_1 < frac_fit_split_human_simple_9)
```

>>>>

---
title: "OptiFit: an improved method for fitting amplicon sequences to existing OTUs"
date: '`r Sys.Date()`'
bibliography: references.bib
link-citations: true
output:
  pdf_document:
    includes:
      in_header: preamble.tex
      before_body: head.tex
    fig_caption: yes
  github_document:
    html_preview: false
csl: msphere.csl
fontsize: 12pt
geometry: margin=1.0in
params:
  include_figures: false
---

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
schtools::set_knitr_opts()
#knitr::opts_knit$set(base.dir = here::here(),
#                     base.url = "../")
knitr::opts_chunk$set(fig.path = 'figures/',
                      dpi = 300)
```

```{r deps, include=FALSE, message = FALSE}
library(assertthat)
library(cowplot)
library(glue)
library(here)
library(knitr)
library(schtools)
library(tidyverse)
set.seed(20210624)

@@

figures_meta <- yaml::read_yaml(here('paper', 'figures.yaml'))

figs <- function(fig_name, attribute, include_figures = params$include_figures) {
  value <- figures_meta[[fig_name]][[attribute]]
  if (attribute == 'cap') {
    value <- glue("\\label{fig:[fig_name]}[value]", .open='[', .close = ']')
  } else if (attribute == 'dim' & !isTRUE(include_figures)) {
    value <- c(0.1,0.1)
  } else {
    value <- eval(parse(text = value))
  }
  return(value)
}

render_figure <- function(fig_name, include_figures = params$include_figures) {
  if (isTRUE(include_figures)) {
    figs(fig_name, 'path') %>%
      tiff::readTIFF() %>%
      grid::grid.raster()
  } else {
    ggplot() + theme_void()
  }
}

@@

Assigning amplicon sequences to operational taxonomic units (OTUs) is an
important step in characterizing microbial communities across large datasets.
OptiClust, a _de novo_ OTU clustering method, produces higher
quality OTUs than other methods at comparable or faster speeds.
A notable difference between _de novo_ clustering and database-dependent
reference clustering methods is that OTU assignments from _de novo_ methods may
change when new sequences are added.
However, one may wish to incorporate new samples to previously clustered
datasets without clustering all sequences again, such as when
comparing across datasets or deploying machine learning models.
Existing reference-based methods produce consistent OTUs, but only consider the
similarity of each query sequence to a single reference sequence in an OTU,
resulting in assignments that are worse than those generated by _de novo_
methods.
To provide an efficient method to fit sequences to existing OTUs, we developed
the OptiFit algorithm.
Like OptiClust, OptiFit considers the similarity of all pairs of reference and
query sequences to produce OTUs of the best quality.
We tested OptiFit using four datasets with two strategies: 1) clustering to a
reference database or 2) splitting the dataset into a reference and query set,
clustering the references using OptiClust, then clustering the queries
to the references.
The result is an improved implementation of reference-based clustering.
OptiFit produces similar quality OTUs as OptiClust at faster speeds when
using the split dataset strategy.
OptiFit provides a suitable option for users requiring consistent OTU
assignments at the same quality afforded by _de novo_ clustering methods.

@@

```{r algorithm, fig.dim=figs('algorithm','dim'), fig.cap=figs('algorithm','cap')}
render_figure('algorithm')
```

@@

```{r workflow, fig.dim=figs('workflow', 'dim'), fig.cap=figs('workflow','cap')}
render_figure('workflow')
```

@@

```{r results_sum, fig.dim=figs('results_sum', 'dim'), fig.cap=figs('results_sum', 'cap')}
render_figure('results_sum')
```

@@

```{r results_split, fig.dim=figs('results_split', 'dim'), fig.cap=figs('results_split', 'cap')}
render_figure('results_split')
```

```{r assert_split_fracs, include = FALSE}
close_enough(cv_fit_split_mcc_human_simple, 0, tol = 10^-3)
assert_that(sec_fit_split_human_simple_1 > sec_fit_split_human_simple_9)
assert_that(frac_fit_split_human_simple_1 < frac_fit_split_human_simple_9)
```
